# Introduction

The purpose of this notebook is to reproduce the analysis by Cosma Shalizi (2011), based on which he concludes that super-linear urban scaling of a set of economic indicators is a statistical artifact. This analysis contradicts previous such claims of super-linear scaling by Bettencourt et al., which have since then grown in popularity, and are now a prominent stylized fact in many theories of economic geography and urban economics.

# Getting the data

The data come from the US Bureau of Economic Analysis (BEA) and a prepared dataset is hosted on the authors homepage.

```{r}
file <- "bea-2006.csv"

if (file.exists(file)) {
  bea <- read.csv(file)
} else {
  bea <- read.csv("http://www.stat.cmu.edu/~cshalizi/mreg/15/lectures/03/bea-2006.csv")
  write.csv(bea, file, row.names = FALSE)
}
```

```{r}
head(bea)
```

The dataset has observations for 366 Metropolitan Statistical Areas (MSAs) in the US for the year 2006, including population size (`pop`), per capita gross metropolitan product (`pcgmp`) and GMP shares across a range of high value-added employment sectors (`finance`, `prof.tech`, `ict`, `management`).

```{r}
dim(bea)
```

There are quite a few missing cases for the latter, though:
```{r}
sum(complete.cases(bea))
```

# Getting to know the data

Here are histograms for all variables:

```{r}
par(mfrow = c(2, 3))
for (col in names(bea)[-1]) {
  hist(bea[,col], main = col, xlab = "")
}
```

From these, we learn that per-capital GMP ranges from 10k USD all the way to 80k USD;

```{r}
bea$MSA[which.min(bea$pcgmp)]
bea$MSA[which.max(bea$pcgmp)]
```

We can also look at pairwise scatterplots:

```{r, fig.width = 10, fig.height = 10}
pairs(bea[,-1], pch = 19, cex = 0.3, log = 2)
```

We see that per capita GMP correlates with population, but so do the shares of GMP in high value-added employment branches, which themselves again correlate with per capita GMP.

# Linear models

Regressing per capita GMP on log population size reveals a reasonable fit:

```{r}
# df <- bea[complete.cases(bea),]
df <- bea
fit <- lm(pcgmp ~ log(pop), data = df)
summary(fit)
plot(pcgmp ~ log(pop), data = df, pch = 19, cex = 0.3)
abline(fit, col = "tomato")
```

This model implies that doubling population comes with a 4500 USD bonus to per capita GMP, roughly speaking.

What is less reasonable is the correlation of residuals with shares in high value added employment sectors:

```{r, fig.width = 10, fig.height = 7}
res <- residuals(fit)
par(mfrow = c(2,2), mar = c(5, 5, 2, 1))
for (col in names(df[, -(1:3)])) {
  y <- df[,col]
  plot(y, res, pch = 19, cex = 0.2, main = NULL, xlab = col, ylab = "Residual")
  lines(smooth.spline(y[!is.na(y)], res[!is.na(y)], df = 4), col = "tomato")
}
```

Controlling for these eliminates the population coefficient almost completely: It went from almost 4500 to 314 and is now eclipsed by its standard error.

```{r}
fit <- lm(
  pcgmp ~ log(pop) + scale(finance) + scale(ict) +
    scale(prof.tech) + scale(management),
  data = bea
)

summary(fit)
```

So basic multiple linear regression is already quite sceptical of the urban scaling effect when employment profiles are taken into account.

We saw quite a few missing cases before, which are currently being dropped. we can also check what happens when we interpolate all missing values by their respective variables' means:


```{r}
bea_interpol <- bea
for (n in names(bea)[-(1:3)]) {
  col <- bea_interpol[,n]
  bea_interpol[is.na(col), n] <- mean(col, na.rm = TRUE)
}

fit <- lm(
  pcgmp ~ log(pop) + scale(finance) + scale(ict) +
    scale(prof.tech) + scale(management),
  data = bea_interpol
)

summary(fit)
```

# Additive models

In the paper, Shalizi (2011) uses generalized additive models (GAMs) to not impose linear relationships. 
