# Introduction

```{r}
library(tinyplot)
tinytheme("ipsum")
```

# Getting the data

First, we read the population data, as obtained from the [OECD data explorer](https://data-explorer.oecd.org/) (population by age and sex).
```{r}
pop <- arrow::read_csv_arrow("data/pop.csv")

# select subset of variables
vars_pop <- c(
  territorial_level = "TERRITORIAL_LEVEL",
  code = "REF_AREA",
  name = "Reference area",
  time = "TIME_PERIOD",
  measure = "MEASURE",
  pop = "OBS_VALUE",
  age = "Age",
  sex = "Sex"
)

pop <- pop[,vars_pop]

# rename variables
names(pop) <- names(vars_pop)

# filter rows to totals and FUAs, drop unused variables
pop <- subset(pop,
  territorial_level == "FUA" & age == "Total" & sex == "Total",
  select = - c(territorial_level, measure, age, sex)
)
```

Then, we do the same with GDP. Here, we need to pay attention to units as there are multiple (percent, per capita, and total) GDP measures.

```{r}
gdp_raw <- arrow::read_csv_arrow("data/gdp.csv")

# select subset of variables
vars_gdp <- c(
  code = "REF_AREA",
  name = "Reference area",
  time = "TIME_PERIOD",
  measure = "MEASURE",
  unit = "UNIT_MEASURE",
  gdp = "OBS_VALUE"
)

gdp <- gdp_raw[,vars_gdp]

# rename variables
names(gdp) <- names(vars_gdp)

# filter rows to GDP, drop unused variables
gdp <- subset(gdp, measure == "GDP", select = -c(measure))

# reshape from long to wide format for different gdp measures
gdp <- reshape(gdp,
  idvar = c("code", "name", "time"),
  timevar = "unit",
  direction = "wide"
)

# rename gdp measures
names(gdp)[4:6] <- c("gdp_pct", "gdp_pc", "gdp")
```

```{r}
pat_raw <- arrow::read_csv_arrow("data/patents.csv")

# select subset of variables
vars_pat <- c(
  code = "REF_AREA",
  name = "Reference area",
  time = "TIME_PERIOD",
  tech = "PATENT_TECH",
  pat = "OBS_VALUE"
)

pat <- pat_raw[,vars_pat]

# rename variables
names(pat) <- names(vars_pat)

# filter rows to pat, drop unused variables
pat <- subset(pat, tech == "_T", select = -c(tech))

```

```{r}
emp <- duckplyr::read_parquet_duckdb("data/local-data-portal/sau_geo_tl_mapping.parquet") |> as.data.frame()
```


Now we combine the datasets by joining on regional codes and time:

```{r}
dat <- merge(gdp, pop, by = c("code", "name", "time"), all = TRUE)
dat <- merge(dat, pat, by = c("code", "name", "time"), all = TRUE)
dat$country <- substr(dat$code, 1, 2)
head(dat)
```

# Getting to know the data

```{r}
pd <- subset(dat, country == "US")
pd <- xtabs(gdp_pc ~ time + code, data = pd)
matplot(rownames(pd), pd, type = "l", lty = 1, xlab = "", ylab = "GDP per capita")
```

# Urban scaling

Here's roughly the analysis that Shalizi (2011) did. First, the original analysis based on extensive data for the US:
```{r}
pd <- subset(dat, country == "US" & time == "2020")
plt(gdp ~ pop, data = pd, cex = 0.5, log = "xy")
```

And then the intensive, per capita, data:

```{r}
pd <- subset(dat, country == "US" & time == "2020")
fit <- lm(gdp_pc ~ log(pop), data = pd)
summary(fit)
plt(gdp_pc ~ log(pop), data = pd, cex = 0.5)
abline(fit, col = "tomato")
```

Just as a point of comparison, let's do the same for a range of countries:

```{r, fig.width = 8, fig.height = 7}

countries <- c("US", "DE", "FR", "IT", "UK", "ES", "PL", "NL", "BE")

pd <- subset(dat, country %in% countries & time == 2020)

tinyplot(
  gdp_pc ~ log(pop) | country,
  data = pd,
  facet = "by", facet.args = list(free = TRUE),
  legend = FALSE, palette = "tableau"
)

tinyplot_add(type = "lm")
```

